---
title: "food_insecurity_analysis"
output: html_document
---
### Load in data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
#install.packages("ggmap", type = "source")
library(ggmap)
```

```{r}
# Read in data
fea <- read_csv('fea_02182017.csv')
colnames(fea)
```

AR and TX are two states with high food insecurity.  Look at these states and examine:
* Note: insecurity values are at the state level
* Which features correlate most with insecurity
* How insecurity and other variables vary at the county level

```{r}
# Subset the data to only AR and/or TX
fea_ARTX <- filter(fea, State == 'AR' | State == 'TX')
fea_AR <- filter(fea, State == 'AR')
fea_TX <- filter(fea, State == 'TX')
```

Plotting on map does not show enough differentiation between FIPS areas

```{r}
library(noncensus)
data("zip_codes")
#zip_codes$fips <- as.character(zip_codes$fips)
head(zip_codes)
fea_AR$FIPS <- as.numeric(fea_AR$FIPS)
fea_AR_geo <- left_join(fea_AR, zip_codes, by = c('FIPS' = 'fips'))

ggplot(fea_AR_geo, aes(x = longitude, y = latitude, fill = GROCPTH12)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar") +
  coord_fixed(1.3)
```

Use histograms to explore county-level data (PCT_LACCESS_POP10, PCT_LACCESS_LOWI10, REDEMP_SNAPS12, FFRPTH12, PCT_NHWHITE10, MEDHHINC10, METRO13, GROCPTH12, SUPERCPTH12)
```{r}
ggplot(fea_ARTX, aes(MEDHHINC10, fill = State)) + geom_histogram()
```

Actually, it's more helpful to compare against the US (**or maybe top insecurity states vs lowest insecurity states**)

Look at food insecurity across time

```{r}
# Could tidy up the data to make this plot better
# Note: jump in time between 00-02 and 07-09
ggplot() + 
  geom_point(data = fea_ARTX, aes(x = State, y = FOODINSEC_00_02), color = 'red') + 
  geom_boxplot(data = fea, aes(x = 'US 00-02', y = FOODINSEC_00_02), color = 'red') + 
  geom_point(data = fea_ARTX, aes(x = State, y = FOODINSEC_07_09), color = 'green') + 
  geom_boxplot(data = fea, aes(x = 'US 07-09', y = FOODINSEC_07_09), color = 'green') + 
  geom_point(data = fea_ARTX, aes(x = State, y = FOODINSEC_10_12), color = 'blue') + 
  geom_boxplot(data = fea, aes(x = 'US 10-12', y = FOODINSEC_10_12), color = 'blue')
```

### Other variables tried out (only showing graphs of interest):
* Some differentiation: MEDHHINC10, POVRATE10, CHILDPOVRATE10, MILK_PRICE10 (sample size too small), FOOD_TAX11, PC_FFRSALES07, PC_FSRSALES07
* No differentiation: PCT_LACCESS_POP10, PCT_LACCESS_LOWI10, REDEMP_SNAPS12, FFRPTH12, FSRPTH12, GROCPTH12, SUPERCPTH12, PCT_65OLDER10, PCT_18YOUNGER10, SNAPSPTH12

```{r}
ggplot() + 
  geom_boxplot(data = fea_ARTX, aes(x = State, y = MEDHHINC10)) + 
  geom_boxplot(data = fea, aes(x = 'US', y = MEDHHINC10))
```

```{r}
ggplot() + 
  geom_boxplot(data = fea_ARTX, aes(x = State, y = POVRATE10)) + 
  geom_boxplot(data = fea, aes(x = 'US', y = POVRATE10))
```

```{r}
ggplot() + 
  geom_boxplot(data = fea_ARTX, aes(x = State, y = CHILDPOVRATE10)) + 
  geom_boxplot(data = fea, aes(x = 'US', y = CHILDPOVRATE10))
```

```{r}
# Can use diverging color scheme / stacked bar chart for above/below
ggplot() + 
  geom_point(data = fea_ARTX, aes(x = State, y = MILK_PRICE10)) + 
  geom_boxplot(data = fea, aes(x = 'US', y = MILK_PRICE10))
```

```{r}
ggplot() + 
  geom_point(data = fea_ARTX, aes(x = State, y = FOOD_TAX11)) + 
  geom_boxplot(data = fea, aes(x = 'US', y = FOOD_TAX11))
```

```{r}
ggplot() + 
  geom_point(data = fea_ARTX, aes(x = State, y = PC_FFRSALES07)) + 
  geom_boxplot(data = fea, aes(x = 'US', y = PC_FFRSALES07))
```

```{r}
ggplot() + 
  geom_point(data = fea_ARTX, aes(x = State, y = PC_FSRSALES07)) + 
  geom_boxplot(data = fea, aes(x = 'US', y = PC_FSRSALES07))
```

### Older stuff

```{r}
file <- '/Users/rachelhz/projects/food_environ_atlas_analysis/food_environment_atlas_2015.xls'
insecurity <- read_excel(file, sheet = 'INSECURITY')
```

```{r}
str(insecurity)
# Plot food insecurity rates (children) by years
ggplot(insecurity) + 
  geom_boxplot(aes(x = 1, y = FOODINSEC_CHILD_01_07)) +
  geom_boxplot(aes(x = 2, y = FOODINSEC_CHILD_03_11))
# Plot food insecurity rates (children) by years against each other
ggplot(insecurity, aes(x = FOODINSEC_CHILD_01_07, y = FOODINSEC_CHILD_03_11)) + geom_point(alpha = 0.3)
```
```{r}
assistance <- read_excel(file, sheet = 'ASSISTANCE')
```

```{r}
head(assistance)
```

### Function for plotting state-level data on maps

For Molly:
* Have ggmap installed and loaded (see top of notebook)
* Use load_state_geo() to bring in state geo data
* Join the state geo data with your data
* Plot the data

```{r}
# Function that loads in state lon/lat data
load_state_geo <- function() {
  # Bring in lat/long data by state name
  state_geo <- map_data("state")
  head(state_geo)
  # Bring in state names and abbreviations
  data(state)
  state_abbrevs <- tibble(abb = state.abb, region = tolower(state.name))
  # Merge to match state abbreviations with lat/long
  state_df <- full_join(state_geo, state_abbrevs, by = "region")
  return(state_df)
}

state_df <- load_state_geo()
head(state_df)
```

```{r}
# DOES NOT WORK, IGNORE!

# Join feature_df into geo data, then plot 'feature_to_plot'
# If you want the joined df, set return_df = TRUE
join_plot_state <- function(state_df, feature_df, feature_to_plot, return_df = FALSE) {
  state_feature <- full_join(state_df, feature_df, by = c('abb' = 'State'))
  
  ggplot(state_feature, aes(x = long, y = lat, fill = feature_to_plot, group = group)) + 
    geom_polygon(color = "white") +
    scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
    coord_fixed(1.3)
  
  #if(return_df == TRUE) {
   # return(state_feature)
  #}
}

join_plot_state(state_df, feature_df, PCT_SNAP14)
```


```{r}
### Here is how to plot by state ###

# Bring in state geo data
state_df <- load_state_geo()

# Summarize 'df' by State using 'group_by_function' on 'feature', then plot by state
# Note: 'df' must have a column of state abbrevs called 'State'

assistance %>% 
  group_by(State) %>% 
  summarise(summ_feature = mean(PCT_SNAP14)) ->
  feature_df

state_feature <- full_join(state_df, feature_df, by = c('abb' = 'State'))

ggplot(state_feature, aes(x = long, y = lat, fill = summ_feature, group = group)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
  coord_fixed(1.3)
```


Use the package 'noncensus' has population and other relevant data for counties and zip codes!!! https://cran.r-project.org/web/packages/noncensus/noncensus.pdf

```{r}
### This didn't work and takes forever! ###
# Try bringing in FIPS (county) level geo data
library(noncensus)

data("zip_codes")
#zip_codes$fips <- as.character(zip_codes$fips)
assistance$FIPS <- as.numeric(assistance$FIPS)
assistance_zip <- full_join(zip_codes, select(assistance, FIPS, State, County, REDEMP_SNAPS12), by = c('fips' = 'FIPS'))

# See if we can plot something by county
# Clearly, we're missing a lot of data
ggplot(assistance_zip, aes(x = longitude, y = latitude, fill = REDEMP_SNAPS12)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
  coord_fixed(1.3)

# Things to do to get per capita data by county, and plot by county
data("counties") # Can use this to get county population!
# 1. Make a FIPS column
counties <- mutate(counties, fips = paste(state_fips, county_fips, sep = ''))
# 2. Create a 'group' column in counties so that plotting happens by county
counties <- mutate(counties, group = 1:nrow(counties))
# 3. Merge counties and assistance data by FIPS
assistance_county <- full_join(assistance, counties, by = c('FIPS' = 'fips'))
# 4. Create new column if need be
assistance_county <- mutate(assistance_county, PCREDEMP_SNAPS12 = REDEMP_SNAPS12 / population)
# (Or plot PC_SNAPBEN08, which has per capita data already and little missing data)
# 5. Merge with zip code lat/lon data
assistance_county_geo <- full_join(assistance_county, zip_codes, by = c('FIPS' = 'fips'))
# 6. Plot!
ggplot(assistance_county_geo, aes(x = longitude, y = latitude, fill = PC_SNAPBEN08, group = group)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
  coord_fixed(1.3)
```

### Compare per capita expenditures on: farm, fast food, restaurants, hh income

```{r}
file <- '/Users/rachelhz/projects/food_environ_atlas_analysis/fea_02182017.csv'
fea <- read_csv(file)
```

Variables of interest:
* PC_DIRSALES07: Direct farm sales per capita, 2007
* PC_FFRSALES07: Expenditures per capita, fast food, 2007 (state level)
* PC_FSRSALES07: Expenditures per capita, restaurants, 2007 (state level)
* MEDHHINC10: Median household income, 2010

```{r}
fea_expend <- select(fea, FIPS, State, PC_DIRSALES07, PC_FFRSALES07, PC_FSRSALES07, MEDHHINC10)
head(fea_expend)
```

```{r}
library(GGally)
# Note: this plot can be misleading because the fast food and restaurant sale data is by state
ggpairs(select(fea_expend, PC_DIRSALES07, PC_FFRSALES07, PC_FSRSALES07, MEDHHINC10), mapping = aes(alpha = 0.5))
```

```{r}
# There are only 50 states
ggplot(fea_expend, aes(PC_FFRSALES07, PC_FSRSALES07)) + geom_point()
```
Overall, full-service restaurant expenditures seem to grow with fast food restaurant expenditures.

```{r}
library(gridExtra)
grid.arrange(
# Fast food expenditures by median hh income
ggplot(fea_expend, aes(MEDHHINC10, PC_FFRSALES07)) + geom_point(size = 0.2, alpha = 0.5), 
# Restaurant expenditures by median hh income
ggplot(fea_expend, aes(MEDHHINC10, PC_FSRSALES07)) + geom_point(size = 0.2, alpha = 0.5), 
nrow = 2)
```
Observations:
* There seems to be a general range ($500-800) of per capita fast food expenditures that applies across the bulk of incomes.
* There seems to be a general range ($400-900) of per capita restaurant expenditures that applies across most incomes.
* Interesting that expenditures for eating out are pretty consistent across rich and poor counties/states.

```{r}
# Plot fast food expenditures on a map

# Bring in state geo data
state_df <- load_state_geo()

# Summarize 'df' by State using 'group_by_function' on 'feature', then plot by state
# Note: 'df' must have a column of state abbrevs called 'State'

fea_expend %>% 
  group_by(State) %>% 
  summarise(PC_FFRSALES07 = mean(PC_FFRSALES07)) ->
  feature_df

state_feature <- full_join(state_df, feature_df, by = c('abb' = 'State'))

ggplot(state_feature, aes(x = long, y = lat, fill = PC_FFRSALES07, group = group)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
  coord_fixed(1.3) + 
  ggtitle('PC_FFRSALES07 by State')
```

```{r}
# Plot restaurant expenditures on a map

# Bring in state geo data
state_df <- load_state_geo()

# Summarize 'df' by State using 'group_by_function' on 'feature', then plot by state
# Note: 'df' must have a column of state abbrevs called 'State'

fea_expend %>% 
  group_by(State) %>% 
  summarise(PC_FSRSALES07 = mean(PC_FSRSALES07)) ->
  feature_df

state_feature <- full_join(state_df, feature_df, by = c('abb' = 'State'))

ggplot(state_feature, aes(x = long, y = lat, fill = PC_FSRSALES07, group = group)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
  coord_fixed(1.3) + 
  ggtitle('PC_FSRSALES07 by State')
```

```{r}
# Plot hh median income on a map

# Bring in state geo data
state_df <- load_state_geo()

# Summarize 'df' by State using 'group_by_function' on 'feature', then plot by state
# Note: 'df' must have a column of state abbrevs called 'State'

fea_expend %>% 
  group_by(State) %>% 
  summarise(MEDHHINC10_mean = mean(MEDHHINC10)) ->
  feature_df

state_feature <- full_join(state_df, feature_df, by = c('abb' = 'State'))

ggplot(state_feature, aes(x = long, y = lat, fill = MEDHHINC10_mean, group = group)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
  coord_fixed(1.3) + 
  ggtitle('MEDHHINC10_mean by State')
```

```{r}
# Plot median hh income by county
data("zip_codes")
zip_codes$fips <- as.character(zip_codes$fips)
joined_zip <- full_join(zip_codes, select(fea_expend, FIPS, MEDHHINC10), by = c('fips' = 'FIPS'))

# See if we can plot something by county
# Clearly, we're missing a lot of data
ggplot(joined_zip, aes(x = longitude, y = latitude, fill = MEDHHINC10, group = fips)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar")+
  coord_fixed(1.3) + 
  ggtitle('MEDHHINC10 by County')
```

```{r}
# Bring in state geo data
state_df <- load_state_geo()

# Summarize 'df' by State using 'group_by_function' on 'feature', then plot by state
# Note: 'df' must have a column of state abbrevs called 'State'

fea_expend %>% 
  group_by(State) %>% 
  summarise(PC_DIRSALES07_mean = mean(PC_DIRSALES07)) ->
  feature_df

state_feature <- full_join(state_df, feature_df, by = c('abb' = 'State'))

ggplot(state_feature, aes(x = long, y = lat, fill = PC_DIRSALES07_mean, group = group)) + 
  geom_polygon(color = "white") +
  scale_fill_continuous(low = "#fee5d9", high = "#a50f15", guide="colorbar") +
  coord_fixed(1.3) + 
  ggtitle('PC_DIRSALES07_mean by State')
```

Observations:
* We have very little farm sales data, so we probably shouldn't use this information.

### Correlation among socioeconomic variables, eg. race against poverty/income/metro, so that we can use a single factor in later analyses

Socioeconomic variables:

```{r, warning=FALSE}
fea_socioecon <- select(fea, FIPS, State, PCT_NHWHITE10, PCT_NHBLACK10, PCT_HISP10, PCT_NHASIAN10, PCT_NHNA10, PCT_NHPI10, PCT_65OLDER10, PCT_18YOUNGER10, MEDHHINC10, POVRATE10, PERPOV10, CHILDPOVRATE10, PERCHLDPOV10, METRO13, POPLOSS00)

# Plot race against income against metro area
ggplot(fea_socioecon, aes(MEDHHINC10, POPLOSS00)) + geom_point(alpha = 0.5) + facet_wrap(~METRO13)
```

Observations (looking at % of category compared to median hh income):
* Out of most races, only Asians seem to appear more in counties with higher median income
* PCT_NHPI10: can remove, make up a low percentage of populations
* PCT_18YOUNGER10: pretty uninformative
* PCT_65OLDER10: not so informative either
* POVRATE10, CHILDPOVRATE10: Poverty rate and child poverty rate are highly linearly related!; as expected, is nearly inversely related to MEDHHINC10
  * most households under the poverty line have children
* PERPOV10: exists for some counties on very low end of MEDHHINC10
* PERCHLDPOV10: exists for a wider range of incomes, compared to PERPOV10, at the low end of the spectrum
 * POPLOSS00: happens more for lower half end of MEDHHINC10

```{r}
ggplot(fea_socioecon, aes(POVRATE10, CHILDPOVRATE10)) + geom_point(alpha = 0.5)
```

